Review on precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-25 manuscript_converted.tex
2022-10-04 conversion from manubot

This fast processing also explains the surprising experiments of fast serial detection which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only 14 ms per image~\citep{ref-EqCCVwl2}.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{figures/visual-latency-estimate.jpg}
\caption{From~\citep{ref-12dWDpY61}}\label{fig:thorpe}
\end{figure}

\subsection{How timing encodes analogous profile}\label{how-timing-encodes-analogous-profile}
An important characteristic of neuronal information is that it consists mainly in the transfer of action potentials, or spikes, which consist of brief impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (that is to say, they are prototypical, all or nothing). An important consequence of the speed of processing is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only 120 ms consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Neurophysiologists tipcally use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{ref-zZUVkTfZ} but also the dependance across neurons~\citep{ref-13T2Gne9z}.

Numerous studies demonstrated the importance of precise timing in neural population activity~\citep{ref-BNK6T5Uk}, efficient encoding thanks to the use of spike latencies~\citep{ref-1DpSM1kpQ,ref-IjepoonT} or precise timing in the auditory system~\citep{ref-mxPMgOmW,ref-1ECJhg2GM}. All these findings, and more~\citep{ref-WZsEyOlK}, highlight the importance of the temporal aspect of the neural code and suggest the existence of repeated spatio-temporal patterns in the spike train. Note also that depending on how an in vitro cell is driven, this may influence the reliability of spike timing in neocortical neurons~\citep{ref-XsHCA4IZ}.

@@

At the level of dynamic processing of visual information, it has been shown that an encoding of the values of luminance imminence in the image instead of the retina~\citep{ref-IjepoonT}. Notably one can appreciate in figure 1 that the response of ganglion cells to visual gratings that are flashed onto the retina. The authors showed that the neuronal response could be encoded in the latency of the response and not only in the frequency of discharge as is often assumed. In figure 4 of the same article, these results are extended to natural images and show a qualitatively similar behavior. The conclusion of the authors is that the discharge latency of the neurons allows to encode spatial characteristics of the image. First-spike latency codes are a feasible mechanism for information transfer even when biologically plausible estimates of stimulus onset are taken into account~\citep{ref-1H1Omh73g}. Interestingly, such fine latency effect can be at the origin of somme visual illusions, for instance the illusion of false colors in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based based on center-surround interactions in the parvocellular pathway~\citep{ref-UlbjMU0W}.

Similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{ref-513xqzmL}. Many models have used these properties in temporal coding to build fast image categorization networks. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{ref-1DpSM1kpQ}. This work has been extended to include unsupervised learning capabilities and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{ref-7K80jcx9,ref-fRudGmR1}. This type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Cameras which, instead of using a basic frame-based representation, uses a representation similar to the one we have just described and which consists in representing the image by events~\citep{ref-Oet1bgdb}. This type of modeling often uses the classical architecture of image categorization developed in deep learning while adapting it to the specificity of the event-based representation~\citep{ref-15cR83gqd}. Note also that timing is not entirely sensorial or internal but in~\citep{ref-1E8gkynCm}, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.'' Altogether, these results show that ``that like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{ref-QK8RfvTZ}.

\subsection{From synfire chains\ldots{}}\label{from-synfire-chains}
The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular~\citep{ref-sYHyiGsP}, which makes it inconsistent with temporal integration of random EPSPs. It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{ref-XsHCA4IZ} (see Figure {[}\ref{fig:mainen}{]}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies on the reliability of spike timing reported in~\citep{ref-1AbvCTG03} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

@@

In his book ``Corticonics''~\citep{ref-19MS6Bbwp}, Moshe Abeles queried if the role of cortical neurons is whether to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electrophysiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{ref-lT1Ato7P}. The book then demonstrates that this could take the form of ``synfire chains,'' that is synchronous activity on subsets of neurons which could be propagated in a stable fashion. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{ref-R2E2wRFg}.

It was shown that a simple model may allow the propagation of such synfire chains~\citep{ref-xbqYi2PD}. This model considers the dynamics of leaky integrate-of-fire neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the weight value, this new activity may get more or less synchronized than the previous pulse (as measured by the standard deviation of spike times in the pulse). Recursively applying this to a sequence of groups generates either a synfire propagation or not. A simple simulation is shown in Figure {[}\ref{fig:diesman}{]}. A crucial aspect of this emergence is in particular the balance between excitation and inhibition~\citep{ref-DL2NpHxr}. This was for instance modeled by feed-forward inhibition which is a fine-scaled latency mechanism that is an essential ingredient in modelling push-pull effects in the primary visual cortex~\citep{ref-mszJ8wdv}.

@@

Note also that synchronicity may explain some unintuitive results. Indeed it has been shown that thalamocortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex as this input is more often synchronously active~\citep{ref-18ZU9zP9l}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{ref-wVZpQvSk}. Interestingly, authors showed that ``Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{ref-1HJjmmSpm} and showed some tuning to to movement direction and reaction time~\citep{ref-FZ9X31Qa}. It is important to note that synchronous events tended to lock to LFP beta waves~\citep{ref-Ejy1sAyh}, and was extended to larger assemblies~\citep{ref-Glu4NlLC} using statistical methods (see~\citep{ref-sec:detection}{\textbf{sec:detection?}}{]}).

In~\citep{ref-3GlNNxH5}, the author proposes a simple spike-based computational framework, based on the idea that stimulus-induced synchrony can be used to extract sensory invariants (for example, the location of a sound source), which is a difficult task for classical neural networks. It relies on the simple remark that a series of repeated coincidences is in itself an invariant. Many aspects of perception rely on extracting invariant features, such as the spatial location of a time-varying sound, the identity of an odor with fluctuating intensity, the pitch of a musical note.

@@

Statistical dependencies in the responses of sensory neurons govern both the amount of stimulus information conveyed and the means by which downstream neurons can extract it. This is put in evidence by analysing the functional significance of correlated firing in a complete population of macaque parasol retinal ganglion cells using a model of multi-neuron spike responses~\citep{ref-Piao2Wzf}. This show precise spatio-temporal differences in this recurrently connected assembly. The different aspects of information in the data are evaluated by a decoding strategy highlighting the role of correlations. A similar dataset used in~\citep{ref-2AsQOOsE} is available from M. Berry's lab~\citep{ref-1FU0Y7Erc}.

@@

More recently Advanced recording techniques have enabled the identification of travelling waves of neuronal activity in different areas of the cortex~\citep{ref-vsKvZJ82}. Authors review these findings, consider the mechanisms by which travelling waves are generated and evaluate their possible roles in cortical function. In particular, spontaneous traveling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states~\citep{ref-BNK6T5Uk}. Studies of sensory-evoked neuronal responses often focus on mean spike rates, with fluctuations treated as internally-generated noise. However, fluctuations of spontaneous activity, often organized as traveling waves, shape stimulus-evoked responses and perceptual sensitivity. The mechanisms underlying these waves are unknown. Further, it is unclear whether waves are consistent with the low rate and weakly correlated ``asynchronous-irregular'' dynamics observed in cortical recordings. Here, authors describe a large-scale computational model with topographically-organized connectivity and conduction delays relevant to biological scales. They find that spontaneous traveling waves are a general property of these networks. The traveling waves that occur in the model are sparse, with only a small fraction of neurons participating in any individual wave. Consequently, they do not induce measurable spike correlations and remain consistent with locally asynchronous irregular states. Further, by modulating local network state, they can shape responses to incoming inputs as observed in vivo.

@@

\subsubsection{Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function~\citep{ref-wVZpQvSk}}\label{spike-synchronization-and-rate-modulation-differentially-involved-in-motor-cortical-function-doi10.1126science.278.5345.1950}
It is now commonly accepted that planning and execution of movements are based on distributed processing by neuronal populations in motor cortical areas~\citep{ref-yLVM9IXi}. It is less clear, though, how these populations organize dynamically to cope with the momentary computational demands. Simultaneously recorded activities of neurons in the primary motor cortex of monkeys during performance of a delayed-pointing task exhibited context-dependent, rapid changes in the patterns of coincident action potentials. Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.

@@

In neuronal models, an efficient use or detection of these spatio-temporal patterns embedded in the spike train comes with the integration of heterogeneous delays~\citep{ref-KS4oxnst,ref-ga3ZTFC7}. Notably, Izhikevich~\citep{ref-SM9G0xBK} introduced the notion of polychronous groups (PGs) as a repetitive motif of spikes defined by a subset of neurons with different, yet precise, spiking delays. This representation has a much greater information capacity in comparison to other neural coding approaches through their connectivity and the possible coexistence of numerous superposed PGs.

@@

\subsection{Are there precise heterosynaptic spiking motifs in the brain?}\label{are-there-precise-heterosynaptic-spiking-motifs-in-the-brain}
Currently, there is a consensus for rate-based coding models in computational and biological neuroscience. Nevertheless, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences (like synfire chains~\citep{ref-Mf6h1KNY}, packets~\citep{ref-2GovoaYw} or hippocampal sequences~\citep{ref-SVfasyCp,ref-PUNGVou} and on the role of such precise spike timing in downstream information transfer and coding~\citep{ref-19fAM6hcv,ref-13dQ0R4UM,ref-MfmcMsqA}. This is for instance relevant in sensory pathways in vision~\citep{ref-1EHZgcKyL}, audition~\citep{ref-FlKJw2VL}, olfaction~\citep{ref-AhSRXBJ2}{]} or touch~\citep{ref-gDinei4a}. In particular, one theoretical viewpoint considers synfire braids (Bienenstock, 1995), where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. In particular, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs (Izhikevich, 2006) which he defined as polychronous groups.

Stereotyped sequences of neuronal activation have been particularly well described in the adult hippocampus and related to its function in mental travel in time and space~\citep{ref-13dQ0R4UM}. These sequences can be internally generated~\citep{ref-SVfasyCp,ref-PUNGVou} and are formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets (Malvache et al., 2016). Thus, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure {[}\ref{fig:haimerl}{]} from~\citep{ref-qI2HUEyy} below). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.

@@

In~\citep{ref-4D8Jnhkz}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a HMM model for quantifying the transitions. ``In this study, van Kempen et al.~show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives interareal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination.''

@@

Such observations suggest that neuronal groups or ensembles, rather than individual neurons, are emergent functional units of cortical activity~\citep{ref-UGHHptIH}. This work shows, using two-photon calcium imaging of populations of neurons from the primary visual cortex of awake mice during visual stimulation and spontaneous activity, that whereas intrinsic ensembles recur at random time intervals, visually evoked ensembles are time-locked to stimuli. It proposes that visual stimuli recruit endogenously generated ensembles to represent visual attributes. Note that evoked ensembles in response to a natural movie played in a loop were precisely timed across repetitions (Fig. 7).

@@

there are temporal delays in the nervous system, both at the neural~\citep{ref-f7gkBktf} and behavioral~\citep{ref-1BSbXPPDF} levels. Extending this knowledge to the optimization of delays in a SNN will provide a breakthrough in the efficiency of these networks.

@@

A popular model for the detection of latency patterns is the tempotron~\citep{ref-1GazUBEdT}. This model is in particular reviewed in~\citep{ref-QCyjy8wE}. The Tempotron is a supervised synaptic learning algorithm which classifies a distractor from a target motif. This aims at extending the perceptron which does not incorporate a spike timing framework. The tempotron learning rule is derived by an optimization process and takes the form of a STDP rule. It is general consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons~\citep{ref-EF6Larp8}. The limits of this model is that it's output is only binary and that it's storage capacity are limited.

Some other models of latency motifs detection using STDP learning rules. For instance,~\citep{ref-Dd3kSDRW} implements a STDP-based spiking deep convolutional neural networks for object recognition;~\citep{ref-vePjgmRh} develops a form of spike-based, competitive learning applied for unsupervised learning. In~\citep{ref-RrgzWOSk}, authors developed novel machine learning tools and statistical tests for unsupervised spatio-temporal pattern detection in non-stationary environments, which was applied to simultaneous electrophysiological recordings from tens to hundreds of neurons for decoding cognitive processes from neural activity.

@@

Memory traces in dynamical systems~\citep{ref-o7rANZux} : ``To perform nontrivial, real-time computations on a sensory input stream, biological systems must retain a short-term memory trace of their recent inputs. It has been proposed that generic high-dimensional dynamical systems could retain a memory trace for past inputs in their current state. This raises important questions about the fundamental limits of such memory traces and the properties required of dynamical systems to achieve these limits. We address these issues by applying Fisher information theory to dynamical systems driven by time-dependent signals corrupted by noise. We introduce the Fisher Memory Curve (FMC) as a measure of the signal-to-noise ratio (SNR) embedded in the dynamical state relative to the input SNR. The integrated FMC indicates the total memory capacity. We apply this theory to linear neuronal networks and show that the capacity of networks with normal connectivity matrices is exactly 1 and that of any network of N neurons is, at most, N. A nonnormal network achieving this bound is subject to stringent design constraints: It must have a hidden feedforward architecture that superlinearly amplifies its input for a time of order N, and the input connectivity must optimally match this architecture. The memory capacity of networks subject to saturating nonlinearities is further limited, and cannot exceed $\sqrt N$. This limit can be realized by feedforward structures with divergent fan out that distributes the signal across neurons, thereby avoiding saturation. We illustrate the generality of the theory by showing that memory in fluid systems can be sustained by transient nonnormal amplification due to convective instability or the onset of turbulence.''

@@

A review of the methods for neuronal response latency estimation including bayesian binning~\citep{ref-Z7LAUup3}.

>>>>

This fast processing also explains the surprising experiments of fast serial detection which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only 14 ms per image~\citep{keysers_speed_2001}.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{figures/visual-latency-estimate.jpg}
\caption{From~\citep{thorpe_seeking_2001}}\label{fig:thorpe}
\end{figure}

\subsection{How timing encodes analogous profile}\label{how-timing-encodes-analogous-profile}
An important characteristic of neuronal information is that it consists mainly in the transfer of action potentials, or spikes, which consist of brief impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (that is to say, they are prototypical, all or nothing). An important consequence of the speed of processing is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only 120 ms consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Neurophysiologists tipcally use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependance across neurons~\citep{ref-13T2Gne9z}.

Numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{ref-1DpSM1kpQ,gollisch_rapid_2008} or precise timing in the auditory system~\citep{ref-mxPMgOmW,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of repeated spatio-temporal patterns in the spike train. Note also that depending on how an in vitro cell is driven, this may influence the reliability of spike timing in neocortical neurons~\citep{mainen_reliability_1995}.

@@

At the level of dynamic processing of visual information, it has been shown that an encoding of the values of luminance imminence in the image instead of the retina~\citep{gollisch_rapid_2008}. Notably one can appreciate in figure 1 that the response of ganglion cells to visual gratings that are flashed onto the retina. The authors showed that the neuronal response could be encoded in the latency of the response and not only in the frequency of discharge as is often assumed. In figure 4 of the same article, these results are extended to natural images and show a qualitatively similar behavior. The conclusion of the authors is that the discharge latency of the neurons allows to encode spatial characteristics of the image. First-spike latency codes are a feasible mechanism for information transfer even when biologically plausible estimates of stimulus onset are taken into account~\citep{chase_first-spike_2007}. Interestingly, such fine latency effect can be at the origin of somme visual illusions, for instance the illusion of false colors in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}.

Similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{ref-513xqzmL}. Many models have used these properties in temporal coding to build fast image categorization networks. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{ref-1DpSM1kpQ}. This work has been extended to include unsupervised learning capabilities and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{ref-7K80jcx9,ref-fRudGmR1}. This type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Cameras which, instead of using a basic frame-based representation, uses a representation similar to the one we have just described and which consists in representing the image by events~\citep{ref-Oet1bgdb}. This type of modeling often uses the classical architecture of image categorization developed in deep learning while adapting it to the specificity of the event-based representation~\citep{ref-15cR83gqd}. Note also that timing is not entirely sensorial or internal but in~\citep{safaie_turning_2020}, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.'' Altogether, these results show that ``that like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.

\subsection{From synfire chains\ldots{}}\label{from-synfire-chains}
The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular~\citep{ref-sYHyiGsP}, which makes it inconsistent with temporal integration of random EPSPs. It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figure {[}\ref{fig:mainen}{]}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies on the reliability of spike timing reported in~\citep{ref-1AbvCTG03} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

@@

In his book ``Corticonics''~\citep{ref-19MS6Bbwp}, Moshe Abeles queried if the role of cortical neurons is whether to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electrophysiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{ref-lT1Ato7P}. The book then demonstrates that this could take the form of ``synfire chains,'' that is synchronous activity on subsets of neurons which could be propagated in a stable fashion. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}.

It was shown that a simple model may allow the propagation of such synfire chains~\citep{ref-xbqYi2PD}. This model considers the dynamics of leaky integrate-of-fire neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the weight value, this new activity may get more or less synchronized than the previous pulse (as measured by the standard deviation of spike times in the pulse). Recursively applying this to a sequence of groups generates either a synfire propagation or not. A simple simulation is shown in Figure {[}\ref{fig:diesman}{]}. A crucial aspect of this emergence is in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This was for instance modeled by feed-forward inhibition which is a fine-scaled latency mechanism that is an essential ingredient in modelling push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}.

@@

Note also that synchronicity may explain some unintuitive results. Indeed it has been shown that thalamocortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex as this input is more often synchronously active~\citep{bruno_cortex_2006}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, authors showed that ``Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{ref-1HJjmmSpm} and showed some tuning to to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tended to lock to LFP beta waves~\citep{denker_lfp_2018}, and was extended to larger assemblies~\citep{ref-Glu4NlLC} using statistical methods (see~\citep{ref-sec:detection}{\textbf{sec:detection?}}{]}).

In~\citep{brette_computing_2012}, the author proposes a simple spike-based computational framework, based on the idea that stimulus-induced synchrony can be used to extract sensory invariants (for example, the location of a sound source), which is a difficult task for classical neural networks. It relies on the simple remark that a series of repeated coincidences is in itself an invariant. Many aspects of perception rely on extracting invariant features, such as the spatial location of a time-varying sound, the identity of an odor with fluctuating intensity, the pitch of a musical note.

@@

Statistical dependencies in the responses of sensory neurons govern both the amount of stimulus information conveyed and the means by which downstream neurons can extract it. This is put in evidence by analysing the functional significance of correlated firing in a complete population of macaque parasol retinal ganglion cells using a model of multi-neuron spike responses~\citep{pillow_spatio-temporal_2008}. This show precise spatio-temporal differences in this recurrently connected assembly. The different aspects of information in the data are evaluated by a decoding strategy highlighting the role of correlations. A similar dataset used in~\citep{schneidman_weak_2006} is available from M. Berry's lab~\citep{ref-1FU0Y7Erc}.

@@

More recently Advanced recording techniques have enabled the identification of travelling waves of neuronal activity in different areas of the cortex~\citep{ref-vsKvZJ82}. Authors review these findings, consider the mechanisms by which travelling waves are generated and evaluate their possible roles in cortical function. In particular, spontaneous traveling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states~\citep{davis_spontaneous_2021}. Studies of sensory-evoked neuronal responses often focus on mean spike rates, with fluctuations treated as internally-generated noise. However, fluctuations of spontaneous activity, often organized as traveling waves, shape stimulus-evoked responses and perceptual sensitivity. The mechanisms underlying these waves are unknown. Further, it is unclear whether waves are consistent with the low rate and weakly correlated ``asynchronous-irregular'' dynamics observed in cortical recordings. Here, authors describe a large-scale computational model with topographically-organized connectivity and conduction delays relevant to biological scales. They find that spontaneous traveling waves are a general property of these networks. The traveling waves that occur in the model are sparse, with only a small fraction of neurons participating in any individual wave. Consequently, they do not induce measurable spike correlations and remain consistent with locally asynchronous irregular states. Further, by modulating local network state, they can shape responses to incoming inputs as observed in vivo.

@@

\subsubsection{Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function~\citep{riehle_spike_1997}}\label{spike-synchronization-and-rate-modulation-differentially-involved-in-motor-cortical-function-doi10.1126science.278.5345.1950}
It is now commonly accepted that planning and execution of movements are based on distributed processing by neuronal populations in motor cortical areas~\citep{grammont_precise_1999}. It is less clear, though, how these populations organize dynamically to cope with the momentary computational demands. Simultaneously recorded activities of neurons in the primary motor cortex of monkeys during performance of a delayed-pointing task exhibited context-dependent, rapid changes in the patterns of coincident action potentials. Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.

@@

In neuronal models, an efficient use or detection of these spatio-temporal patterns embedded in the spike train comes with the integration of heterogeneous delays~\citep{ref-KS4oxnst,zhang_supervised_2020}. Notably, Izhikevich~\citep{ref-SM9G0xBK} introduced the notion of polychronous groups (PGs) as a repetitive motif of spikes defined by a subset of neurons with different, yet precise, spiking delays. This representation has a much greater information capacity in comparison to other neural coding approaches through their connectivity and the possible coexistence of numerous superposed PGs.

@@

\subsection{Are there precise heterosynaptic spiking motifs in the brain?}\label{are-there-precise-heterosynaptic-spiking-motifs-in-the-brain}
Currently, there is a consensus for rate-based coding models in computational and biological neuroscience. Nevertheless, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences (like synfire chains~\citep{ref-Mf6h1KNY}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{ref-SVfasyCp,ref-PUNGVou} and on the role of such precise spike timing in downstream information transfer and coding~\citep{ref-19fAM6hcv,ref-13dQ0R4UM,ref-MfmcMsqA}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996}{]} or touch~\citep{johansson_first_2004}. In particular, one theoretical viewpoint considers synfire braids (Bienenstock, 1995), where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. In particular, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs (Izhikevich, 2006) which he defined as polychronous groups.

Stereotyped sequences of neuronal activation have been particularly well described in the adult hippocampus and related to its function in mental travel in time and space~\citep{ref-13dQ0R4UM}. These sequences can be internally generated~\citep{ref-SVfasyCp,ref-PUNGVou} and are formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets (Malvache et al., 2016). Thus, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure {[}\ref{fig:haimerl}{]} from~\citep{haimerl_internal_2019} below). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.

@@

In~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a HMM model for quantifying the transitions. ``In this study, van Kempen et al.~show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives interareal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination.''

@@

Such observations suggest that neuronal groups or ensembles, rather than individual neurons, are emergent functional units of cortical activity~\citep{miller_visual_2014}. This work shows, using two-photon calcium imaging of populations of neurons from the primary visual cortex of awake mice during visual stimulation and spontaneous activity, that whereas intrinsic ensembles recur at random time intervals, visually evoked ensembles are time-locked to stimuli. It proposes that visual stimuli recruit endogenously generated ensembles to represent visual attributes. Note that evoked ensembles in response to a natural movie played in a loop were precisely timed across repetitions (Fig. 7).

@@

there are temporal delays in the nervous system, both at the neural~\citep{perrinet_active_2014} and behavioral~\citep{khoei_flash-lag_2017} levels. Extending this knowledge to the optimization of delays in a SNN will provide a breakthrough in the efficiency of these networks.

@@

A popular model for the detection of latency patterns is the tempotron~\citep{ref-1GazUBEdT}. This model is in particular reviewed in~\citep{gutig_spike_2014}. The Tempotron is a supervised synaptic learning algorithm which classifies a distractor from a target motif. This aims at extending the perceptron which does not incorporate a spike timing framework. The tempotron learning rule is derived by an optimization process and takes the form of a STDP rule. It is general consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons~\citep{caporale_spike_2008}. The limits of this model is that it's output is only binary and that it's storage capacity are limited.

Some other models of latency motifs detection using STDP learning rules. For instance,~\citep{ref-Dd3kSDRW} implements a STDP-based spiking deep convolutional neural networks for object recognition;~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning. In~\citep{ref-RrgzWOSk}, authors developed novel machine learning tools and statistical tests for unsupervised spatio-temporal pattern detection in non-stationary environments, which was applied to simultaneous electrophysiological recordings from tens to hundreds of neurons for decoding cognitive processes from neural activity.

@@

Memory traces in dynamical systems~\citep{ganguli_memory_2008} : ``To perform nontrivial, real-time computations on a sensory input stream, biological systems must retain a short-term memory trace of their recent inputs. It has been proposed that generic high-dimensional dynamical systems could retain a memory trace for past inputs in their current state. This raises important questions about the fundamental limits of such memory traces and the properties required of dynamical systems to achieve these limits. We address these issues by applying Fisher information theory to dynamical systems driven by time-dependent signals corrupted by noise. We introduce the Fisher Memory Curve (FMC) as a measure of the signal-to-noise ratio (SNR) embedded in the dynamical state relative to the input SNR. The integrated FMC indicates the total memory capacity. We apply this theory to linear neuronal networks and show that the capacity of networks with normal connectivity matrices is exactly 1 and that of any network of N neurons is, at most, N. A nonnormal network achieving this bound is subject to stringent design constraints: It must have a hidden feedforward architecture that superlinearly amplifies its input for a time of order N, and the input connectivity must optimally match this architecture. The memory capacity of networks subject to saturating nonlinearities is further limited, and cannot exceed $\sqrt N$. This limit can be realized by feedforward structures with divergent fan out that distributes the signal across neurons, thereby avoiding saturation. We illustrate the generality of the theory by showing that memory in fluid systems can be sustained by transient nonnormal amplification due to convective instability or the onset of turbulence.''

@@

A review of the methods for neuronal response latency estimation including bayesian binning~\citep{levakova_review_2015}.

\subsection{decoding neural activity}\label{decoding-neural-activity}
In generic linear non linear lnl models, the output is assumed to be poisson. As such a simple decoding strategy is to asscume it is to b inferned for a given tuning curves (Jayazeri) or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes such that the decoding amounts to a single-layer logistic regression.

Unitary event analysis is performed by a statistical model of coincidence detection~\citep{ref-13iJohyPk}. This was extensively used in detecting above chance significant synchronous patterns, in particular in recordings of pairs of neurons (see~\citep{riehle_spike_1997} for instance).
